Cluster Dynamics with HETDEX at z < 0.5 - I: Simulated Performance, Mass Recovery, and Limits to Cosmology

### 2017-02-01 paper1.tex
a few more edits and a figure tweak

To test for this, we used the Millennium simulation \citep{Springel2005a} as an alternative data set. The Millennium simulation adopts a flat cosmological model based of the values derived from the Two-degree Field Galaxy Redshift Survey \citep{Colless2001} and the first year data of the \emph{Wilkinson Microwave Anisotropy Probe} (WMAP; \citealt{Spergel2003}): $\Omega_\Lambda = 0.75$, $\Omega_M = 0.25$, $\sigma_8 = 0.9$, $n_s = 1$ and $H_0= 73$ \kms \mpc. As this cosmology differs dramatically from that assumed in the Buzzard catalogs (see Section~\ref{sec: Buzzard}, the clusters in Millennium provide an independent mock sample to understand how a training sample derived from Buzzard will impact the mass recovery on a wholly new dataset.

We repeat our analysis using cluster halo and \cite{DeLucia2007} galaxy catalogs derived from the Millennium simulation and obtained via querying the Millennium online database\footnote{\url{http://gavo.mpa-garching.mpg.de/Millennium/}} \citep{Lemson2006}. The Millennium simulation tracks $2160^3$ dark matter particles of $8.6\times 10^8 ~h^{-1}$ \Msol\ inside a comoving 500 $(h^{-1} \mathrm{Mpc})^3$ box from $z=127$ to 0. The following SQL query selects the cluster halos:
\begin{lstlisting}[language=SQL]
SELECT *
FROM   mfield..fof
WHERE  snapnum IN (SELECT snapnum
                   FROM   snapshots..mr
                   WHERE  z BETWEEN 0.01 AND 0.5)
       AND m_crit200 > 100
       AND random BETWEEN 0 AND 10000
\end{lstlisting}
To select the corresponding galaxies, we use:
\begin{lstlisting}[language=SQL, breaklines=true]
SELECT sh.fofid,
       sh.subhaloid,
       g.galaxyid,
       g.x,
       g.y,
       g.z,
       g.redshift,
       g.velx,
       g.vely,
       g.velz,
       sdss.g_sdss,
       sdss.r_sdss,
       sdss.i_sdss
FROM   mfield..fofsubhalo AS sh
       INNER JOIN mpagalaxies..delucia2006a AS g
              ON sh.subhaloid = g.subhaloid
       INNER JOIN mpagalaxies..delucia2006a_sdss2mass AS sdss
              ON g.galaxyid = sdss.galaxyid
WHERE  sh.fofid IN (SELECT fofid
       FROM   mfield..fof
       WHERE  snapnum IN (SELECT snapnum
                      FROM   snapshots..mr
                      WHERE  z BETWEEN 0.01 AND 0.5)
              AND m_crit200 > 1000
              AND random BETWEEN 0 AND 10000)
\end{lstlisting}

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/millBuzzComparison.pdf}
	\caption[ML based cluster mass predictions in the Millennium Simulation]{\emph{Top:} ML based cluster mass predictions for the Millennium simulation clusters where the ML method has been trained with either a subset of the Millennium clusters (solid line) or the Buzzard catalog (dashed line). The shaded areas show the 68\% scatter around the median. The solid black line shows the 1:1 relation. \emph{Bottom:} The fractional error (Equation~\ref{eq: fractional error}) also as a function of true cluster mass. The similarity of the predictions with the different training sets demonstrates how the ML method is not sensitive to the underlying cosmological assumptions.} \label{fig: mill buzz comparison}
\end{figure}

We conduct our test in two ways. Both use the ML methods (see Section \ref{sec:machine learning method}) to predict the cluster masses of the Millennium clusters, but each test uses a different training set. First, we use the full set of clusters detected in the Buzzard catalogs (14,000 clusters with $M > 10^{11}$ \msol) to train the ML. Second, the Millennium clusters are split into training-testing samples. This provides a test case where we have different cosmological choices between the training and testing samples (Buzzard training; Millennium testing), and the same cosmological assumptions in both training and testing. In addition, it provides two different treatments of galaxy evolution in cluster environments. The differing galaxy evolution effects can impact observed galaxy magnitude and color evolution. This provides an important check to understand whether the specific assumptions made in the models affect the results of this analysis.

The top panel of Figure~\ref{fig: mill buzz comparison} shows the ML predicted cluster masses for the $\sim$4000 Millennium clusters as a function of true cluster mass. The orange (Millennium) and green (Buzzard) colors indicate the two different training samples. The median (solid and dashed lines) predicted cluster masses show similar trends regardless of the training data set used. The bottom panel of Figure~\ref{fig: mill buzz comparison} shows the fraction error (Equation~\ref{eq: fractional error}) also as a function of true cluster mass. The large amount of scatter (the shaded area) in the fractional error for the Buzzard-trained predictions is due to the training set including clusters with masses below the $M = 10^{13}$ \Msol\ threshold for the Millennium clusters. This allows the ML method to predict masses which can be significantly different, whereas the Millennium training set does not include $M < 10^{13}$ \Msol\ clusters, which artificially reduces the scatter of the predicted masses.

@@

For a survey such as HETDEX, we can estimate the constraints on $\sigma_8$ using Equation \ref{eq:sigElapprox}. If we consider clusters with masses above $10^{14}$ \Msol\ and with Perfect knowledge observations, the lowest mass scale uncertainty (given in Table~\ref{tbl:mass scatter}) is $\Delta_{log_{10}}M \sim 0.075$ dex, corresponding to fractional uncertainty of $\approx20\%$. This gives a uncertainty on $\sigma_8$ of 7\%. For clusters above $10^{14}$ \Msol, Survey observations constrain the masses to about 51\% which, in turn, constrains $\sigma_8$ to 20\%.

Because we assume $\Omega_m$ is perfectly known from other data, and the superior quality of the data (no contamination, signal-to-noise issues, etc.), realistic expectations for HETDEX is to directly constrain $\sigma_8$ is not yet competitive with other methods (\eg, CMB, WL, X-ray). For example, \cite{DeHaan2016} constrain $\sigma_8$ to $\sim5$\% using a sample of 337 SZE detected clusters from the SPT-SZE survey.

>>>>

To test for this, we used the Millennium simulation \citep{Springel2005a} as an alternative data set. The Millennium simulation tracks $2160^3$ dark matter particles of $8.6\times 10^8 ~h^{-1}$ \Msol\ inside a comoving 500 $(h^{-1} \mathrm{Mpc})^3$ box from $z=127$ to 0. It adopts a flat cosmological model based of the values derived from the Two-degree Field Galaxy Redshift Survey \citep{Colless2001} and the first year data of the \emph{Wilkinson Microwave Anisotropy Probe} (WMAP; \citealt{Spergel2003}): $\Omega_\Lambda = 0.75$, $\Omega_M = 0.25$, $\sigma_8 = 0.9$, $n_s = 1$ and $H_0= 73$ \kms \mpc. As this cosmology differs dramatically from that assumed in the Buzzard catalogs (see Section~\ref{sec: Buzzard}, the clusters in Millennium provide an independent mock sample to understand how a training sample derived from Buzzard will impact the mass recovery on a wholly new dataset.

We repeat our analysis using cluster halo and \cite{DeLucia2007} galaxy catalogs obtained via querying the Millennium on-line database\footnote{\url{http://gavo.mpa-garching.mpg.de/Millennium/}} \citep{Lemson2006}. The following SQL query selects the cluster halos:
\begin{lstlisting}[language=SQL]
SELECT *
FROM   mfield..fof
WHERE  snapnum IN (SELECT snapnum
                   FROM   snapshots..mr
                   WHERE  z BETWEEN 0.01 AND 0.5)
       AND m_crit200 > 100
       AND random BETWEEN 0 AND 10000
\end{lstlisting}
To select the corresponding galaxies, we use:
\begin{lstlisting}[language=SQL, breaklines=true]
SELECT sh.fofid,
       sh.subhaloid,
       g.galaxyid,
       g.x,
       g.y,
       g.z,
       g.redshift,
       g.velx,
       g.vely,
       g.velz,
       sdss.g_sdss,
       sdss.r_sdss,
       sdss.i_sdss
FROM   mfield..fofsubhalo AS sh
       INNER JOIN mpagalaxies..delucia2006a AS g
              ON sh.subhaloid = g.subhaloid
       INNER JOIN mpagalaxies..delucia2006a_sdss2mass AS sdss
              ON g.galaxyid = sdss.galaxyid
WHERE  sh.fofid IN (SELECT fofid
       FROM   mfield..fof
       WHERE  snapnum IN (SELECT snapnum
                      FROM   snapshots..mr
                      WHERE  z BETWEEN 0.01 AND 0.5)
              AND m_crit200 > 1000
              AND random BETWEEN 0 AND 10000)
\end{lstlisting}

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/millBuzzComparison.pdf}
	\caption[ML based cluster mass predictions in the Millennium Simulation]{\emph{Top:} ML based cluster mass predictions for the clusters selected from the Millennium database where the ML method has been trained with either a subset of the Millennium clusters (solid line) or the Buzzard catalog (dashed line). The shaded areas show the 68\% scatter around the median. The solid black line shows the 1:1 relation. \emph{Bottom:} The fractional error (Equation~\ref{eq: fractional error}) also as a function of true cluster mass. The similarity of the predictions with different training sets demonstrates how the ML method is not sensitive to the underlying cosmological assumptions.} \label{fig: mill buzz comparison}
\end{figure}

We conduct our test in two ways. Both use the ML methods (see Section \ref{sec:machine learning method}) to predict the cluster masses of the Millennium clusters, but each test uses a different training set. First, we use the full set of clusters detected in the Buzzard catalogs (14,000 clusters with $M > 10^{11}$ \msol) to train the ML. This provides a test case where we have different cosmological choices between the training and testing samples (Buzzard training; Millennium testing). Second, the Millennium clusters are split into training-testing samples, which provides a benchmark to test against where the same cosmological assumptions are made in both training and testing. In addition, use of the Millennium simulation provides two different treatments of galaxy evolution in cluster environments. The differing galaxy evolution effects can impact observed galaxy magnitude and color evolution. This provides an important check to understand whether the specific assumptions made in the models affect the results of this analysis.

The top panel of Figure~\ref{fig: mill buzz comparison} shows the ML predicted cluster masses for the $\sim$4000 Millennium clusters as a function of true cluster mass. The orange (Millennium) and green (Buzzard) colors indicate the two different training samples. The median (solid and dashed lines) predicted cluster masses show similar trends regardless of the training data set used. A simple $\chi^2$-test indicates the two median lines are indistinguishable from one another, $\chi^2=0.004$. The bottom panel of Figure~\ref{fig: mill buzz comparison} shows the fraction error (Equation~\ref{eq: fractional error}) also as a function of true cluster mass. The large amount of scatter (the shaded area) in the fractional error for the Buzzard-trained predictions is due to the training set including clusters with masses below the $M = 10^{13}$ \Msol\ threshold for the Millennium clusters. This allows the ML method to predict masses which can be significantly different, whereas the Millennium training set does not include $M < 10^{13}$ \Msol\ clusters, which artificially reduces the scatter of predicted masses. The ML method trained on Buzzard clusters has a mean scatter increase of $\sim0.05$ dex, a fractional increase of about 12\%, across all true cluster masses.

@@

For a survey such as HETDEX, we can estimate the constraints on $\sigma_8$ using Equation \ref{eq:sigElapprox}. If we consider clusters with masses above $10^{14}$ \Msol\ and with Perfect knowledge observations, the lowest mass scale uncertainty (given in Table~\ref{tbl:mass scatter}) is $\Delta_{log_{10}}M \sim 0.075$ dex, corresponding to fractional uncertainty of $\sim20\%$. This gives an uncertainty on $\sigma_8$ of 7\%. For clusters above $10^{14}$ \Msol, Survey observations constrain the masses to about 51\% which, in turn, constrains $\sigma_8$ to 20\%.

Because we assume $\Omega_m$ is perfectly known from other data, and the superior quality of the data (no contamination, signal-to-noise issues, etc.), realistic expectations for HETDEX is to directly constrain $\sigma_8$ is not yet competitive with other methods (\eg, CMB, WL, X-ray). For example, \cite{DeHaan2016} constrain $\sigma_8$ to $\sim5$\% using a sample of 337 SZE detected clusters from the SPT-SZE survey, compared to the 7\% uncertainty with HETDEX and Perfect observations.
